\large{\bf Looking forwards and backwards through locally regulated populations }

### 2023-04-12 forwards-and-backwards.tex
clumpy figure
apx start

%%%%%%% notation
\newcommand{\DG}{\mathcal{B}}  % generator of the dispersal process
\newcommand{\DD}{\mathcal{D}}  % the second-order part of the generator of dispersal
\newcommand{\meanq}{\vec b}    % mean of dispersal, times theta
\newcommand{\covq}{\mathbf{C}}     % covariance matrix of dispersal, times theta
\newcommand{\kernel}{\rho}  % interaction kernels
\newcommand{\smooth}[1]{\kernel_{#1} \! * \!}  % convolution by the interaction kernel
\newcommand{\wavespeed}{\mathfrak{c}}    % speed (vector) of a wave
\newcommand{\Lgen}{\mathcal{L}}    % generator of a lineage
\newcommand{\Pgen}{\mathcal{P}}    % generator of the population process
\newcommand{\lp}{\xi}              % process with levels
\newcommand{\labelspace}{\mathcal{I}} % space of labels
\newcommand{\concat}{\oplus}   % concatenation of labels
\newcommand{\measures}{\mathcal{M}_F(\IR^d)} % finite measures on Rd
\newcommand{\cmeasures}{\mathcal{M}_F(\overline{\IR}^d)} % finite measures on compactified Rd
\newcommand{\lpmeasures}{\mathcal{M}(\overline{\IR^d} \times [0,\infty))} % locally finite measures on space x level

@@

\begin{figure}
    \begin{center}
        \comment{TODO} FIGURE
    \end{center}
    \caption{
        Two panels: left is 2D picture of clumped population;
        right is a bumpy expanding wavefront.
        \label{fig:clumping}
    }
\end{figure}

@@

\begin{lemma}
    \label{def: MP definition of limit}
For $G \in \mathcal{C}^{\infty}(\IR)$ with $\|G'''\|_\infty<\infty$,
and 
$f \in \mathcal{C}_{b}^\infty(\overline{\IR}^d)$, define the 
function $G_f$ by 
$G_f(\eta):=G (\langle f, \eta \rangle)$.
Let $\Pgen^N$ be the generator given by
\begin{equation}
\begin{aligned}
    \Pgen^N G_f(\eta)
    &:=
    G'(\langle f, \eta \rangle)
        \theta \bigg\langle
        \gamma(x, \eta)
            \int f(z) r(z,\eta) q_\theta(x,dz)
        -
        f(x)\mu_\theta(x, \eta),
        \eta(dx) \bigg\rangle
    \\&\qquad {}
    + \frac{1}{2} G''(\langle f, \eta \rangle)
    \frac{\theta}{N}
    \bigg\langle
        \gamma(x, \eta) \int f^2(y) r(y, \eta) q_\theta(x, dy) + f^2(x) \mu_\theta(x, \eta),
    \eta(dx) \bigg\rangle .
\end{aligned}
\end{equation}
A process $(\eta^N_t)_{t \geq 0}$ 
satisfies the martingale characterization of Definition~\ref{defn:mgale_construction}
if and only if it is a solution 
to the $(\Pgen^N, \eta_0)$-martingale problem, i.e.,
if for all such test functions 
$$
    M_t :=
    G_f(\eta^N_t) - G_f(\eta^N_0)
    - \int_{0}^{t}\Pgen^N G_f(\eta^N_s)ds
$$
is a martingale (with respect to the natural $\sigma$-field).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results on Lookdown Representations}
    \label{sec: Markov Mapping Theorem Application}

>>>>

%%%%%%% notation
\newcommand{\DG}{\mathcal{B}}  % generator of the dispersal process
\newcommand{\DD}{\mathcal{D}}  % the second-order part of the generator of dispersal
\newcommand{\meanq}{\vec b}    % mean of dispersal, times theta
\newcommand{\covq}{\mathbf{C}}     % covariance matrix of dispersal, times theta
\newcommand{\kernel}{\rho}  % interaction kernels
\newcommand{\smooth}[1]{\kernel_{#1} \! * \!}  % convolution by the interaction kernel
\newcommand{\wavespeed}{\mathfrak{c}}    % speed (vector) of a wave
\newcommand{\Lgen}{\mathcal{L}}    % generator of a lineage
\newcommand{\Pgen}{\mathcal{P}}    % generator of the population process
\newcommand{\lp}{\xi}              % process with levels
\newcommand{\labelspace}{\mathcal{I}} % space of labels
\newcommand{\concat}{\oplus}   % concatenation of labels
\newcommand{\measures}{\mathcal{M}_F(\IR^d)} % finite measures on Rd
\newcommand{\cmeasures}{\mathcal{M}_F(\overline{\IR}^d)} % finite measures on compactified Rd
\newcommand{\lpmeasures}{\mathcal{M}(\overline{\IR^d} \times [0,\infty))} % locally finite measures on (compactified space) x level
\newcommand{\flpmeasures}{\mathcal{M}(\IR^d \times [0,\infty))} % locally finite measures on space x level

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=0.45\textwidth]{figures/clumpy_2d/clumpy_123.density.3.0}
        \comment{TODO} FIGURE
    \end{center}
    \caption{
        Two panels: left is 2D picture of clumped population;
        right is a bumpy expanding wavefront.
        \label{fig:clumping}
    }
\end{figure}

@@

\comment{TODO: the previous definition of $\Pgen^N$ was the following wrong thing; fixup below}
\begin{equation}
\begin{aligned}
    \text{ONLY RIGHT TO $O(1/N)$} \qquad
    \Pgen^N G_f(\eta)
    &:=
    G'(\langle f, \eta \rangle)
        \theta \bigg\langle
        \gamma(x, \eta)
            \int f(z) r(z,\eta) q_\theta(x,dz)
        -
        f(x)\mu_\theta(x, \eta),
        \eta(dx) \bigg\rangle
    \\&\qquad {}
    + \frac{1}{2} G''(\langle f, \eta \rangle)
    \frac{\theta}{N}
    \bigg\langle
        \gamma(x, \eta) \int f^2(y) r(y, \eta) q_\theta(x, dy) + f^2(x) \mu_\theta(x, \eta),
    \eta(dx) \bigg\rangle .
\end{aligned}
\end{equation}

\begin{lemma}
    \label{def: MP definition of limit}
For $G \in \mathcal{C}^{\infty}(\IR)$ with $\|G'''\|_\infty<\infty$,
and 
$f \in \mathcal{C}_{b}^\infty(\overline{\IR}^d)$, define the 
function $G_f$ by 
$G_f(\eta):=G (\langle f, \eta \rangle)$.
Let $\Pgen^N$ be the generator given by
\begin{equation}
\begin{aligned}
    \Pgen^N G_f(\eta)
    &:=
    \theta \bigg\langle
    \gamma(x, \eta)
        \int 
        \left(G(\langle f, \eta \rangle + f(z)/N) - G(\langle f, \eta \rangle)\right)
        r(z,\eta) q_\theta(x,dz)
    +
    \left(G(\langle f, \eta \rangle - f(x)/N) - G(\langle f, \eta \rangle)\right)
    \mu_\theta(x, \eta),
    \eta(dx) \bigg\rangle .
\end{aligned}
\end{equation}
A process $(\eta^N_t)_{t \geq 0}$ 
satisfies the martingale characterization of Definition~\ref{defn:mgale_construction}
if and only if it is a solution 
to the $(\Pgen^N, \eta_0)$-martingale problem, i.e.,
if for all such test functions 
$$
    M_t :=
    G_f(\eta^N_t) - G_f(\eta^N_0)
    - \int_{0}^{t}\Pgen^N G_f(\eta^N_s)ds
$$
is a martingale (with respect to the natural $\sigma$-field).

@@

% %% %% %% %% %% %% %% %% %% %% %% %% %% %% %% %% %%
\section{Markov Mapping Theorem}

% %% %% %% %% %% %% %% %% %% %% %% %% %% %% %% %% %%
\subsection{Lookdown Generators}

In this section we verify one of the conditions of the Markov Mapping Theorem,
namely, that ``integrating out levels'' in the generator of the lookdown process
we obtain the generator of the projected process.
We will work with test functions of the form
$$  f(\lp) = \prod_{(x, u) \in \lp} g(x, u) = \exp\left(\langle \log g, \lp \rangle \right) . $$
Furthermore, recall that $\kappa^N(\lp)(\cdot) = \lp(\cdot \times [0, N)) / N$
is the ``spatial projection operator'',
and define the probability kernel $\Gamma^N : \measures \to \mathcal{M}(\IR^d \times [0,N))$
so that for $\eta \in \measures$, if $\hat g(x) = \int_0^N g(x, u) du / N$, then
\begin{align*}
    F^N_g(\eta)
    &:=
    \int f(\lp) \Gamma^N(\eta, d\lp) \\
    &=
    \exp\left(
        N \left\langle
            \log \frac{1}{N} \int_0^N g(x, u) du, \eta(dx)
        \right\rangle
    \right) \\
    &=
    \exp\left(
        N \left\langle \log \hat g(x) \eta(dx) \right\rangle
    \right) ,
\end{align*}
i.e., $\Gamma^N$ assigns independent labels on $[0, N]$ to each of the points in $\eta$.
It follows from Lemma~\ref{def: MP definition of limit} that for test functions of this form
\comment{(beware the different definition of $f$ to that lemma)}
the generator of $\eta^N_t$ is
\begin{align}
    \Pgen^N F^N_g(\eta)
    &=
    F^N_g(\eta)
    \theta \bigg\langle
        \gamma(x, \eta) \int r(z, \eta) \left( \hat g(z) - 1 \right) q_\theta(x, dz)
        +
        \mu_\theta(x, \eta) \left( \frac{1}{\hat g(x)} - 1 \right)
        ,
        \eta(dx)
    \bigg\rangle
\end{align}
The generator of $\lp^N_t$ is $A^N$, defined in equation~\label{eqn:lookdown_generator}.

\begin{lemma}
    \label{lem:averaged_generators}
    For all finite counting measures $\eta$ on $\IR^d$,
    \begin{align}
        \int A^N f(\lp) \Gamma^N(\eta, d\lp)
        &=
        \Pgen^N F^N_g(\eta) .
    \end{align}
\end{lemma}

For the limiting process,
we now require test functions such that $g(x,u) = 1$ for all $u$ larger than some $u_0$;
As before, recall that $\kappa(\lp)(\cdot) = \lim_{u \to \infty} \lp(\cdot \times [0, u)) / u$
is the ``spatial projection operator'',
and define the probability kernel $\Gamma : \measures \to \flpmeasures$
so that for $\eta \in \measures$,
defining $h(x) = \int_0^\infty (g(x, u) - 1) du$,
\begin{align*}
    F_g(\eta)
    &:=
    \int f(\lp) \Gamma(\eta, d\lp) \\
    &=
    \exp\left(
        \big \langle
            \int_0^\infty (g(x, u) - 1) du,
            \eta(dx)
        \big \rangle
    \right) \\
    &=
    e^{ \langle h(x), \eta(dx) \rangle } .
\end{align*}
i.e., $\Gamma(\eta, \cdot)$ is the distribution of a conditionally Poisson process
with intensity a product of $\eta$ and Lebesgue measure.
It again follows from Lemma~\ref{def: MP definition of limit} that for test functions of this form
\comment{(beware the different definition of $f$ to that lemma)}
the generator of $\eta_t$ is
\begin{align}
    \Pgen^\infty F_g(\eta)
    &=
    F_g(\eta)
    \left\langle
        \gamma(x, \eta)
        \DG( h(\cdot) r(\cdot) )(x)
        + F(x, \eta) h(x)
        + \alpha \gamma(x, \eta) r(x, \eta) h^2(x)
        ,
        \eta(dx)
    \right\rangle .
\end{align}
The generator of $\lp_t$ is $A$, defined in equation~\eqref{eqn:limiting_lookdown_generator},

\begin{lemma}
    \label{lem:averaged_generators_limit}
    For all $\eta \in \measures$,
    \begin{align}
        \int A f(\lp) \Gamma(\eta, d\lp)
        &=
        \Pgen^\infty F_g(\eta) .
    \end{align}
\end{lemma}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{OLDER STUFF Results on Lookdown Representations}
    \label{sec: Markov Mapping Theorem Application}
